Exploring probability distributions for bivariate temporal granularities

### 2019-08-05 main.Rmd
spell check and mixed single gran intro

\begin{equation} \label{eq1}
\begin{split}
f_{(x,y)}(z) & = f_{(x, x+1)}(z) + c(x, x+1)(f_{(x+1, y)}(z)-1) \\  
  & =  f(x, x+1) + c(x, x+1)[f(x+1, x+2) + c(x+1, x+2)(f(x+2, y) - 1) - 1]\\
  & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+1)c(x+1, x+2)(f(x+2, y) - 1)\\
  & =  f(x, x+1) + c(x, x+1)(f(x+1, x+2) - 1) + c(x, x+2)(f(x+2, y) - 1)\\
  &\vdots\\
  & = \sum_{i=0}^{order(y) - order(x) - 1} c(x, x+i)(f(x
  +i, x+i+1) - 1)\\
\end{split}
\end{equation}

To elaborate more, let us consider the following hierarchy table.

@@

From Equation(1), we have
\begin{equation} \label{eq1}
\begin{split}
f_(hhour, week)(z) = f_(hhour, hour)(z) + c(hhour, hour)f_(hour,day) + c(hhour, day)f(day, week)\\
               = hhour_hour + 2hour_day + 2*24day_week
\end{split}
\end{equation}

@@

This is valid in a scenario when the relationship between x and (x+1) is periodic (Define periodic in Definition section). However, when x and x+1 is not periodic, then some modification is required in Equation(1).

Consider, a hierarchy table of the form below:

```{r, hierarchy_tbl}
hierarchy_model <- tibble::tibble(units = c("minute", "hour", "day", "month", "quarter", "semester", "year"), conversion_fac = c(60, 24, NA, 3, 2, 2, 1))
knitr::kable(hierarchy_model)
```

The table suggests that one hour is composed of 60 minutes, 1 day is composed of 24 hours, 1 quarter is composed of 3 months, 1 semester is consists of 2 quarters and 1 year is made up of 2 semesters. These partitions are periodic, since the number of finer units within each of these coarses units are same always.

@@

Suppose, we are not given the intermediate single-order-up granularities but still want all possible multiple order-up granularties. In that case, it is necessary to atleast have the have the functional form that relates the smallest temporal unit to the highest temporal unit. In that case, we are able to form single-order-up from this functional relationship and still use the recursive relation to get the multiple order-up granularities.

@@

Another example could be where $C_1$ maps row numbers to Day-of-Week and $C_2$ maps row numbers to Month-of-Year. Here $C_1$ can take 7 values while $C_2$ can take 12 values. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of DOW and MOY. All of these are non-empty because every DOW can occur in every month. So graphics involving $C_1$ and $C_2$ aare potentially useful.

# Visualisation

<!-- Why visualise -->
Analysts often want to fit their data to statistical models, either to test hypotheses or predict future values. However, improper choice of models can lead to wrong predictions. One important use of visualization is exploratory data analysis, which is gaining insight into how data is distributed to inform data transformation and modeling decisions.

@@

<!-- Why distribution plot -->
<!-- Huge data difficult to summarise using one statistic-->
But with huge amount of data being available, sometimes mean, median or any one summary statistic is not enough to understand a dataset. Soon enough following questions become more interesting:

@@

 - Does values rise very quickly between 25th percentile and median but not as quickly between median and 75th percentile? More generally, how the variation is the dataset changes across different percentiles/deciles?

@@

The entire distribution can be visualized or some contextual summary statistics can be visualised to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anolmalies in the dataset.

@@

Moreover, much like the quartiles divide the dataset equally into four equal parts, extensions might include dividing the dataset even further. The deciles plots consist of 9 values that split the dataset into ten parts and the percentile plot consists of 99 values that split the dataset into hundred parts. A large dataset is required before the extreme percentiles can be estimated with any accuracy.

>>>>

\begin{equation} \label{eq1}
\begin{split}
f_{(x,y)}(z) & = f_{(x, x+1)}(z) + c(x, x+1)(f_{(x+1, y)}(z)-1) \\  
  & =  f_{(x, x+1)}(z) + c(x, x+1)[f_{(x+1, x+2)}(z) + c(x+1, x+2)(f_{(x+2, y)}(z) - 1) - 1]\\
  & =  f_{(x, x+1)}(z) + c(x, x+1)(f_{(x+1, x+2)}(z) - 1) + c(x, x+1)c(x+1, x+2)(f_{(x+2, y)}(z) - 1)\\
  & =  f_{(x, x+1)}(z) + c(x, x+1)(f_{(x+1, x+2)}(z) - 1) + c(x, x+2)(f_{(x+2, y)}(z) - 1)\\
  &\vdots\\
  & = \sum_{i=0}^{order(y) - order(x) - 1} c(x, x+i)(f_{(x
  +i, x+i+1)}(z) - 1)\\
\end{split}
\end{equation}

For illustration, let us consider the following hierarchy table.

@@

From Equation(1), we have
\begin{equation} \label{eq1}
\begin{split}
f_{(hhour, week)}(z) & = f_{(hhour, hour)}(z) + c(hhour, hour)f_{(hour,day)}(z) + c(hhour, day)f_{(day, week)}(z) \\
              & = {hhour\_hour}(z) + 2{hour\_day}(z) + 2*24{day\_week}(z) \\
\end{split}
\end{equation}

@@

<!-- This is valid in a scenario when the relationship between x and (x+1) is periodic (Define periodic in Definition section). However, when x and x+1 is not periodic, then some modification is required in Equation(1). -->

Suppose we extend the hierarchy table to include the linear granularities month and quarter. Since months consists of unequal number of days, any linear granularity which is higher in order than months will also have unequal number of days. Hence, any calendar categorization which includes one linear granularity with orders at most as days and another one with orders at least as months will be aperiodic in nature. However, it is very much possible for a calendar categorization consisting of two linear granularities higher in order than month to be periodic.For example, the categorization month_quarter would be periodic since each quarter always consists of three months.

```{r table_accessor2}
kable(tibble::tibble(
  `linear granularities`  = c("second", "minute", "hhour", "hour", "day", "week", "month", "quarter"),
  `Conversion factor` = c(60, 30, 2, 24, 7, "aperiodic", 3, 1), 
  `Single-order-up accessors` = c("second_minute", "minute_hhour", "hhour_hour", "hour_day", "day_week", "week_month", "month_quarter", 1)), format = "markdown")
```

The table suggests that one hour is composed of 60 minutes, 1 day is composed of 24 hours, 1 quarter is composed of 3 months, 1 semester is consists of 2 quarters and 1 year is made up of 2 semesters. These partitions are periodic, since the number of finer units within each of these coarser units are same always.

@@

Suppose, we are not given the intermediate single-order-up granularities but still want all possible multiple order-up granularties. In that case, it is necessary to at least have the have the functional form that relates the smallest temporal unit to the highest temporal unit. In that case, we are able to form single-order-up from this functional relationship and still use the recursive relation to get the multiple order-up granularities.

@@

Another example could be where $C_1$ maps row numbers to Day-of-Week and $C_2$ maps row numbers to Month-of-Year. Here $C_1$ can take 7 values while $C_2$ can take 12 values. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of DOW and MOY. All of these are non-empty because every DOW can occur in every month. So graphics involving $C_1$ and $C_2$ are potentially useful.

# Visualization

<!-- Why visualize -->
Analysts often want to fit their data to statistical models, either to test hypotheses or predict future values. However, improper choice of models can lead to wrong predictions. One important use of visualization is exploratory data analysis, which is gaining insight into how data is distributed to inform data transformation and modeling decisions.

@@

<!-- Why distribution plot -->
<!-- Huge data difficult to summaries using one statistic-->
But with huge amount of data being available, sometimes mean, median or any one summary statistic is not enough to understand a data set. Soon enough following questions become more interesting:

@@

 - Does values rise very quickly between 25th percentile and median but not as quickly between median and 75th percentile? More generally, how the variation is the data set changes across different percentiles/deciles?

@@

The entire distribution can be visualized or some contextual summary statistics can be visualized to emphasize certain properties of the distribution. These properties can throw light on central tendency, skewness, kurtosis, variation of the distribution and can also be useful in detecting extreme behavior or anomalies in the data set.

@@

Moreover, much like the quartiles divide the data set equally into four equal parts, extensions might include dividing the data set even further. The deciles plots consist of 9 values that split the data set into ten parts and the percentile plot consists of 99 values that split the data set into hundred parts. A large data set is required before the extreme percentiles can be estimated with any accuracy.
